Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-12-30 analysis/paper/paper.qmd
updates

However, a few Multifacial cores and Polyhedrons also appeared in the near-spherical region. This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded shapes and led to their grouping alongside spheroids. Some artifacts are identified as spheroids or subspheroids but grouped as ellipsoid or rounded box, because they exhibit a long axis. This reflects the use of different classification criteria in typological and quantitative approaches: typological assessments consider artifacts close to spherical, particularly when they exhibit large edge angles and rounded surfaces, whereas spherical harmonic analysis emphasizes aspects such as overall symmetry.

@@

Diacritical analysis of spheroid and subspheroid surfaces reveals a structured and recurrent reduction sequence comprising three phases (@fig-spheroids-2797-1098). Phase one was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation. Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface. Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form. Phase two was peripheral centripetal exploitation. In this phase, previous removals in phase one serve as platforms for centripetal flaking around the core periphery. This represents a surface exploitation stage. A notable feature is the angle between the face's ventral surface and platform (i.e. core edge angle), which often exceeds 90°. This phase contributes significantly to the increasingly spherical shape of the artifact. Phase three was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing roundness.

@@

# 计算凸包函数
get_hull <- function(df) {
  if(nrow(df) < 3) return(df)  
  hull_idx <- chull(df$Z, df$IS)
  df[hull_idx, ]
}

# 为每个类型生成凸包
hull_data <- benn.output %>%
  group_by(type) %>%
  do(get_hull(.)) %>%
  ungroup()

# 定义颜色映射
type_colors <- c(
  "multifacial" = "#b2abd2",
  "polyhedron" = "#5e3c99", 
  "subspheroid" = "#fdb863",
  "spheroid" = "#e66101"
)

@@

# 绘图
p_benn_ternary <- ggtern(data = benn.output, aes(x = Z, y = IS, z = EL)) +
  # 添加三角网格掩膜
  geom_mask() +

  # 凸包填充 - 添加在半透明层
  geom_polygon(
    data = hull_data,
    aes(fill = type, group = type),
    color = NA,  # 无边框
    alpha = 0.2  # 填充透明度
  ) +

  # 凸包边框
  geom_polygon(
    data = hull_data,
    aes(color = type, group = type),
    fill = NA,
    size = 0.8,
    alpha = 0.8
  ) +

  # 散点
  geom_point(
    aes(color = type),
    size = 3,
    alpha = 0.8
  ) +

  # 主题和标签
  theme_bw() +
  theme_showarrows() +
  theme_clockwise() +

  # 轴标签
  labs(
    x = "Planar",
    xarrow = "Planarity",
    y = "Isotropic", 
    yarrow = "Isotropy",
    z = "Linear",
    zarrow = "Elongation",
    color = "Type",
    fill = "Type"
  ) +

  # 颜色映射
  scale_color_manual(
    values = type_colors,
    limits = names(type_colors)
  ) +

@@

# 保存图形
output_dir <- here("analysis", "paper", "figures")
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

@@

# 在文档中显示图形
knitr::include_graphics(file.path(output_dir, "benn_ternary_plot.jpg"))

@@

# 保存组合图像
output_dir <- here("analysis", "paper", "figures")
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

@@

# 在文档中显示图形
knitr::include_graphics(here("analysis/paper/figures/combined_spearman_correlation_matrices.png"))

>>>>

However, a few Multifacial cores and Polyhedrons also appeared in the near-spherical region. This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded shapes and led to their grouping alongside spheroids. Some artifacts are identified as spheroids or subspheroids but grouped as ellipsoid or rounded box, because they exhibit a long axis. This reflects the use of different classification criteria in typological and quantitative approaches: typological assessments consider artifacts close to spherical, particularly when they exhibit large edge angles and rounded surfaces, whereas spherical harmonic analysis emphasizes aspects such as absolute symmetry.

@@

Diacritical analysis of spheroid and subspheroid surfaces reveals a structured and recurrent reduction sequence comprising three phases (@fig-spheroids-2797-1098). Phase one was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation. Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface. Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form. Phase two was peripheral centripetal exploitation. In this phase, previous removals in phase one serve as platforms for centripetal flaking around the core periphery. This represents a surface exploitation stage. A notable feature is the angle between the flaking surface and platform (i.e. core edge angle), which often exceeds 90°. This phase contributes significantly to the increasingly spherical shape of the artifact. Phase three was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing roundness.

@@

get_hull <- function(df) {
  if(nrow(df) < 3) return(df)  
  hull_idx <- chull(df$Z, df$IS)
  df[hull_idx, ]
}

hull_data <- benn.output %>%
  group_by(type) %>%
  do(get_hull(.)) %>%
  ungroup()

type_colors <- c(
  "multifacial" = "#b2abd2",
  "polyhedron" = "#5e3c99", 
  "subspheroid" = "#fdb863",
  "spheroid" = "#e66101"
)

@@

# 绘图
p_benn_ternary <- ggtern(data = benn.output, aes(x = Z, y = IS, z = EL)) +

  geom_mask() +

  geom_polygon(
    data = hull_data,
    aes(fill = type, group = type),
    color = NA,
    alpha = 0.2
  ) +

  geom_polygon(
    data = hull_data,
    aes(color = type, group = type),
    fill = NA,
    size = 0.8,
    alpha = 0.8
  ) +

  geom_point(
    aes(color = type),
    size = 3,
    alpha = 0.8
  ) +

  theme_bw() +
  theme_showarrows() +
  theme_clockwise() +

  labs(
    x = "Planar",
    xarrow = "Planarity",
    y = "Isotropic", 
    yarrow = "Isotropy",
    z = "Linear",
    zarrow = "Elongation",
    color = "Type",
    fill = "Type"
  ) +

  scale_color_manual(
    values = type_colors,
    limits = names(type_colors)
  ) +

@@

output_dir <- here("analysis", "paper", "figures")
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

@@

knitr::include_graphics(file.path(output_dir, "benn_ternary_plot.jpg"))

@@

output_dir <- here("analysis", "paper", "figures")
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

@@

knitr::include_graphics(here("analysis/paper/figures/combined_spearman_correlation_matrices.png"))
